---
title: "Setup: install everything at once"
---

```{mermaid}
flowchart LR
    A[Install R ≥ 4.3<br/>+ RStudio/Positron] --> B[Run setup/install.R<br/>one command, all R packages]
    B --> C{Python weeks?}
    C -- yes --> D[conda env create<br/>-f setup/environment.yml]
    C -- no --> E[Verify install]
    D --> E
```

## 1. R and an IDE

- Install **R ≥ 4.3** from [CRAN](https://cran.r-project.org/)
- Install [RStudio Desktop](https://posit.co/download/rstudio-desktop/) (or [Positron](https://positron.posit.co/))
- Install [Quarto](https://quarto.org/docs/get-started/) (bundled with recent RStudio)

## 2. All R packages in one command

From the repo root:

```{r}
#| eval: false
source("setup/install.R")
```

This uses **pak**, which resolves CRAN, Bioconductor, and GitHub dependencies together and installs in parallel — typically 10–15 minutes on a fresh machine.

## 3. Python environment (Weeks 4, 5, 8)

Optional but recommended. Install [Miniforge](https://github.com/conda-forge/miniforge), then:

```bash
conda env create -f setup/environment.yml
conda activate biostats-course
```

## 4. Verify your installation

```{r}
#| eval: false
pkgs <- c("sommer","lme4","mixOmics","BGLR","tidymodels","xgboost",
          "DESeq2","shiny","plumber","agridat")
ok <- sapply(pkgs, requireNamespace, quietly = TRUE)
data.frame(package = pkgs, installed = ok)
```

All `TRUE`? You're ready for [Week 0](weeks/week0a-biostats-primer.qmd).

::: {.callout-tip}
## Reproducibility
The repo ships an `renv.lock` snapshot. To reproduce exact versions later: `renv::restore()`.
:::
