---
title: "Week 0a — Biostatistics Primer"
subtitle: "The concepts every analysis rests on"
---

```{mermaid}
flowchart LR
    A[Populations<br/>vs samples] --> B[Distributions<br/>normal · binomial · Poisson · NB]
    B --> C[Hypothesis testing<br/>p-values · errors · power]
    C --> D[Confidence<br/>intervals]
    D --> E[Experimental design<br/>replication · randomization · blocking]
    E --> F[Multiple testing<br/>preview for GWAS]
```

## What biostatistics is

Biostatistics is the science of learning from biological data in the presence of **variation**. Two plants of the same genotype never yield identically — the job is to separate *signal* (genetics, treatment) from *noise* (environment, measurement) and to state honestly how sure we are.

## Populations, samples, and estimation

- **Population**: every canola plant that could ever be grown from a cultivar.
- **Sample**: the 200 plots you actually harvested.
- **Parameter** (unknown truth, e.g. true mean yield μ) vs **estimate** (your sample mean x̄).

Statistical inference quantifies how far x̄ may sit from μ.

## Distributions you will meet constantly

```{r}
#| code-fold: true
#| fig-height: 3.2
library(tidyverse)
set.seed(1)
d <- bind_rows(
  tibble(x = rnorm(4000, 40, 6),        dist = "Normal — yield (bu/ac)"),
  tibble(x = rbinom(4000, 10, .3),      dist = "Binomial — diseased plants /10"),
  tibble(x = rnbinom(4000, mu = 20, size = 2), dist = "Neg. binomial — RNA-seq counts")
)
ggplot(d, aes(x)) + geom_histogram(bins = 40, fill = "#159957") +
  facet_wrap(~dist, scales = "free") + theme_minimal()
```

| Distribution | Crop example | Model family |
|---|---|---|
| Normal | yield, height, oil % | linear model |
| Binomial | disease incidence | logistic GLM |
| Poisson / Negative binomial | counts; RNA-seq reads | count GLM (NB handles overdispersion) |

## Hypothesis testing in five sentences

1. State a **null** H₀ (e.g., "no yield difference between lines").
2. Compute a test statistic from the data.
3. The **p-value** = probability of a statistic at least this extreme *if H₀ were true*. It is **not** the probability H₀ is true.
4. **Type I error** (α): rejecting a true H₀ — false positive. **Type II** (β): missing a real effect. **Power** = 1 − β, driven by replication, effect size, and noise.
5. A tiny p with a trivial effect size is not a finding — always report effect sizes.

```{r}
t.test(extra ~ group, data = sleep)   # classic built-in example
```

## Confidence intervals

A 95% CI is the set of parameter values compatible with the data: across repeated experiments, 95% of such intervals capture the truth. Prefer CIs to bare p-values — they show magnitude *and* uncertainty.

## Experimental design — where good statistics starts

- **Replication** — quantifies noise; no reps, no inference.
- **Randomization** — protects against hidden bias (soil gradients).
- **Blocking** — put comparisons close together: RCBD, alpha-lattice, row–column designs used in every serious yield trial. Week 1 models these directly.

## Why multiple testing matters (GWAS preview)

Test 1 SNP at α = 0.05 → 5% false-positive risk. Test 500,000 SNPs → ~25,000 false positives expected under the null. Week 2–3 pages use Bonferroni and FDR to fix this.

## Exercises

1. Simulate 1,000 t-tests under the null (`rnorm` both groups). What fraction have p < 0.05? Why?
2. For a trait with σ = 5 and a true difference of 2, use `power.t.test()` to find n per group for 80% power.
3. Sketch (on paper) an RCBD for 20 lines, 3 blocks, on a field with a known fertility gradient.
