---
title: "Week 6 — Interactive Web Applications with Shiny"
subtitle: "From analysis to a deployed tool collaborators can click"
---

```{mermaid}
flowchart LR
    A[UI + server<br/>anatomy] --> B[Reactivity<br/>dependency graph]
    B --> C[Manhattan + GS<br/>explorer app]
    C --> D[Deploy<br/>shinyapps.io]
    D --> E[Share URL<br/>portfolio evidence]
```

## 1. Anatomy of a Shiny app

Two objects: a **UI** (layout + inputs/outputs) and a **server** (how outputs depend on inputs). Reactivity is a dependency graph — change `input$trait` and only downstream pieces recompute.

Three functions cover most needs: `reactive()` (shared computed value), `render*()` (outputs), `observeEvent()` (side-effects like downloads).

## 2. The course app: GWAS + GS explorer

Full source in [`shiny-app/app.R`](https://github.com/sampathperumal/genomic-biostats-course/blob/main/shiny-app/app.R). Core skeleton:

```{r}
#| eval: false
library(shiny); library(bslib); library(plotly); library(DT)

gwas <- readRDS("gwas_results.rds")   # trait, chr, pos, pvalue (Weeks 2–4 output)

ui <- page_sidebar(
  title = "Crop GWAS & GS Explorer — Sampath Perumal",
  sidebar = sidebar(
    selectInput("trait", "Trait", unique(gwas$trait)),
    sliderInput("thr", "-log10(p) threshold", 2, 10, 5, step = .5)
  ),
  card(card_header("Manhattan plot"), plotlyOutput("man")),
  card(card_header("Significant markers"), DTOutput("tbl"))
)

server <- function(input, output, session) {
  gw <- reactive(dplyr::filter(gwas, trait == input$trait))
  output$man <- renderPlotly({
    d <- gw()
    plot_ly(d, x = ~cum_pos, y = ~-log10(pvalue),
            color = ~factor(chr %% 2), colors = c("#159957", "#1e6bb8"),
            type = "scattergl", mode = "markers",
            text = ~marker, hoverinfo = "text+y") |>
      layout(showlegend = FALSE,
             shapes = list(list(type = "line", x0 = 0, x1 = max(d$cum_pos),
                                y0 = input$thr, y1 = input$thr,
                                line = list(dash = "dot"))))
  })
  output$tbl <- renderDT(dplyr::filter(gw(), -log10(pvalue) > input$thr))
}

shinyApp(ui, server)
```

## 3. Reactivity rules that prevent 90% of bugs

- Compute once in a `reactive()`, use it in many outputs — never repeat filtering inside each `render*()`.
- No side-effects inside `reactive()`; those belong in `observeEvent()`.
- Big data: load once outside `server`, filter inside.

## 4. Deploy in 10 minutes

```{r}
#| eval: false
install.packages("rsconnect")
rsconnect::setAccountInfo(name = "ACCOUNT", token = "TOKEN", secret = "SECRET")
rsconnect::deployApp("shiny-app/")
```

Free tier at [shinyapps.io](https://www.shinyapps.io) → a public URL. Put that URL on your CV and in job applications — a *deployed* app is proof, a script is a claim.

## 5. Level-ups

`golem` for packaged production apps · `shinytest2` for testing · Docker for self-hosting · connect to Week 7's database instead of RDS files.

## Exercises

1. Run the app locally with Week 3/4 outputs saved as `gwas_results.rds`.
2. Add a GS tab: observed-vs-predicted scatter with a model selector (ridge/BayesB/RF).
3. Add `downloadHandler` to export the filtered marker table as CSV.
4. Deploy and send the URL to a colleague on their phone.
