---
title: "Week 1 — Linear & Mixed Models"
subtitle: "From ANOVA to spatial mixed models on real prairie trials"
---

```{mermaid}
flowchart LR
    A[Linear model<br/>y = Xβ + e] --> B[GLM<br/>logistic · counts]
    B --> C[Mixed model<br/>y = Xβ + Zu + e]
    C --> D[BLUE vs BLUP<br/>shrinkage & h²]
    D --> E[Spatial models<br/>AR1×AR1 on field trials]
```

**Dataset:** `agridat::gilmour.serpentine` — a real South Australian wheat trial (Gilmour et al. 1997), the canonical spatial-analysis teaching set, plus Yates' oats for the split-plot classic.

## 1. The linear model

```{r}
library(tidyverse); library(agridat); library(lme4); library(emmeans)
data(oats)
lm1 <- lm(yield ~ gen + factor(nitro), data = oats)
anova(lm1)
```

Fixed effects, F-tests — the machinery beneath every ANOVA table you've published.

## 2. GLMs in one chunk

```{r}
# Binomial GLM: disease incidence example (simulated scores on real structure)
set.seed(2)
oats$diseased <- rbinom(nrow(oats), 10, plogis(-1 + 0.02 * scale(oats$yield)))
glm1 <- glm(cbind(diseased, 10 - diseased) ~ gen, family = binomial, data = oats)
summary(glm1)$coefficients |> head(3)
```

The same idea powers RNA-seq DE (negative-binomial GLM — Week 5).

## 3. The mixed model: y = Xβ + Zu + e

Blocks and genotypes become **random effects**: they share a distribution, so estimates *borrow strength* and shrink toward the mean.

```{r}
mm1 <- lmer(yield ~ factor(nitro) + (1 | gen) + (1 | block), data = oats)
summary(mm1)$varcor
```

**Heritability from variance components** (line-mean basis):

```{r}
vc <- as.data.frame(VarCorr(mm1))
Vg <- vc$vcov[vc$grp == "gen"]; Ve <- vc$vcov[vc$grp == "Residual"]
h2 <- Vg / (Vg + Ve / 3)        # 3 blocks
round(h2, 2)
```

**BLUP shrinkage, visualized:**

```{r}
#| fig-height: 3
blup <- ranef(mm1)$gen[,1]
blue <- coef(lm(yield ~ gen - 1, oats)) - mean(oats$yield)
tibble(line = names(blup), BLUE = as.numeric(blue), BLUP = blup) |>
  pivot_longer(-line) |>
  ggplot(aes(value, line, colour = name)) + geom_point(size = 3) +
  labs(x = "Deviation from mean", colour = NULL) + theme_minimal()
```

BLUPs sit closer to zero — shrinkage in proportion to reliability. This *is* what genomic BLUP does with markers.

## 4. Spatial analysis of a real field trial

```{r}
library(sommer)
data(gilmour.serpentine)
g <- gilmour.serpentine |> mutate(R = as.numeric(row), C = as.numeric(col))

m_base <- mmer(yield ~ 1, random = ~ gen, data = g, verbose = FALSE)
m_spat <- mmer(yield ~ 1,
               random = ~ gen + spl2Da(C, R),
               data = g, verbose = FALSE)
c(base = m_base$AIC, spatial = m_spat$AIC)
```

The spatial spline soaks up field trend; genotype variance is estimated more cleanly and rankings change — decisions change.

```{r}
#| fig-height: 3.2
ggplot(g, aes(C, R, fill = yield)) + geom_tile() +
  scale_fill_viridis_c() + labs(title = "Raw field trend — why we model space") +
  theme_minimal()
```

## Interview-grade summary

- Random vs fixed: *"Random effects share a distribution; I model genotypes as random to get shrunken BLUPs and variance components for heritability."*
- Spatial: *"AR1×AR1 or 2-D splines model field trend; ignoring it inflates error and mis-ranks lines."*

## Exercises

1. Refit `m_spat` with row and column as additional random factors; compare AIC.
2. Compute h² from `m_spat` variance components (`vpredict`).
3. In Yates' oats, treat nitrogen as numeric and fit a quadratic response — is the linear model adequate?

**Data credits:** Gilmour et al. (1997) *JABES*; Yates (1935), via the `agridat` package (Kevin Wright).
