---
title: "Week 5 — RNA-seq & Multi-Omics Integration"
subtitle: "Negative-binomial GLMs, networks, and factor models — canola-flavoured"
---

```{mermaid}
flowchart LR
    A[Counts → NB GLM<br/>DESeq2] --> B[Multiple testing<br/>FDR]
    B --> C[WGCNA<br/>modules & eigengenes]
    C --> D[MOFA2<br/>factors across omics]
    D --> E[Biomarkers &<br/>interpretation]
```

**Data:** a downsampled real *Brassica napus* RNA-seq count matrix (via `data/get_data.R`, pulled from NCBI GEO) with a simulated-but-realistic methylation layer for the integration demo — swap in your own matrices with two lines.

## 1. Why counts need negative binomial

RNA-seq counts are overdispersed (variance > mean): Poisson understates noise → false positives. DESeq2 models counts ~ NB, shrinks per-gene dispersions toward a trend (**empirical Bayes** — Week 3's idea reappearing), and tests log-fold-changes with a Wald or LRT test. The design formula is Week 1's linear model: `~ batch + condition`.

```{r}
#| eval: false
library(DESeq2)
counts <- readRDS("data/bnapus_counts.rds")     # genes x samples
meta   <- readRDS("data/bnapus_meta.rds")       # condition, batch
dds <- DESeqDataSetFromMatrix(counts, meta, design = ~ batch + condition)
dds <- DESeq(dds)
res <- results(dds, alpha = 0.05)
summary(res)
```

```{r}
# Self-contained demo so the page always renders: simulate NB counts w/ signal
library(tidyverse)
set.seed(9)
n_g <- 2000; de <- 1:150
mu  <- rexp(n_g, 1/200)
cts <- sapply(1:6, \(i) rnbinom(n_g, mu = mu * ifelse(i > 3 & seq_len(n_g) %in% de, 2, 1), size = 3))
pv  <- apply(cts, 1, \(x) wilcox.test(x[1:3], x[4:6])$p.value)
padj <- p.adjust(pv, "BH")
tibble(gene = 1:n_g, padj, truly_DE = gene %in% de) |>
  count(sig = padj < .05, truly_DE)
```

**FDR (Benjamini–Hochberg)** controls the *proportion* of false discoveries — with 20k genes, the right currency. Bonferroni controls any single false positive and is appropriate for GWAS-style confirmatory claims.

## 2. Co-expression: WGCNA in concept

Correlation matrix → soft-threshold power (approximate scale-free topology) → topological overlap → hierarchical modules → **module eigengene** (first PC of the module) → correlate eigengenes with traits. It's Week 2's PCA working inside a network.

```{r}
#| eval: false
library(WGCNA)
sft <- pickSoftThreshold(t(vst_counts))
net <- blockwiseModules(t(vst_counts), power = sft$powerEstimate, minModuleSize = 30)
moduleTraitCor <- cor(net$MEs, traits)
```

## 3. MOFA2: one factor model across omics layers

MOFA2 = factor analysis across matched RNA-seq + methylation (+ genotype PAV) matrices: latent **factors** capture shared variation, with per-layer variance-explained telling you which omics drives which factor. Unsupervised, handles missing layers, the modern default.

```{r}
#| eval: false
library(MOFA2)
mofa <- create_mofa(list(RNA = rna_mat, Methylation = meth_mat))
mofa <- prepare_mofa(mofa, model_options = get_default_model_options(mofa))
mofa <- run_mofa(mofa, use_basilisk = TRUE)
plot_variance_explained(mofa)
plot_top_weights(mofa, view = "RNA", factor = 1, nfeatures = 15)
```

Supervised counterpart: **DIABLO** (`mixOmics::block.splsda`) when you have groups to discriminate and want a cross-omics biomarker panel.

## 4. The three questions to ask any omics analysis

1. Is the **design confounded** (batch ⟂ condition)? Check before modeling; fix with design terms or ComBat/RUV.
2. Is the **test matched to the distribution** (NB for counts, logit-M for methylation proportions)?
3. Is **multiplicity controlled**, and is FDR the right currency for the claim?

## Exercises

1. Run `data/get_data.R` to fetch the canola GEO matrix; repeat the DESeq2 block for real; volcano-plot the results.
2. In the simulation, raise replicates from 3→6 per group: how do power and the false-discovery count change?
3. Build a 2-layer MOFA on the provided matrices; which layer dominates Factor 1?

**Credits:** DESeq2 (Love et al. 2014), WGCNA (Langfelder & Horvath 2008), MOFA2 (Argelaguet et al. 2020).
