---
title: "Week 2 — Multivariate Statistics & Dimension Reduction"
subtitle: "PCA, PLS-DA, and clustering on crop diversity data"
---

```{mermaid}
flowchart LR
    A[Scaling &<br/>distance] --> B[PCA<br/>structure & QC]
    B --> C[Clustering<br/>k-means · hierarchical · mclust]
    C --> D[sPLS-DA<br/>supervised, sparse]
    D --> E[Validation<br/>permutation · CV]
```

**Dataset:** `agridat::gauch.soy` multi-environment soybean for PCA-of-GxE, plus a built-in genomic marker matrix from `sommer::DT_cpdata` (real 363-line apple-progeny SNP data) standing in for a pea/canola panel — swap in your own VCF-derived matrix with one line.

## 1. PCA — what it actually does

PCA eigen-decomposes the covariance matrix: PC1 is the direction of maximum variance, PC2 the next, orthogonal. On a **marker matrix**, leading PCs approximate ancestry — the population structure you correct for in GWAS.

```{r}
library(tidyverse); library(sommer); library(factoextra)
data(DT_cpdata)                 # DT = pheno, GT = markers (-1/0/1)
M <- GT_cpdata
p <- prcomp(M, scale. = TRUE)
fviz_eig(p, addlabels = TRUE, ncp = 8)
```

```{r}
#| fig-height: 3.5
scores <- as_tibble(p$x[,1:2]) |> mutate(color = DT_cpdata$color)
ggplot(scores, aes(PC1, PC2, colour = color)) + geom_point(alpha = .7) +
  theme_minimal() + labs(title = "Genetic structure from SNPs")
```

::: {.callout-tip}
**Swap in your crop:** `M <- vcfR::extract.gt(...)` → numeric matrix → identical workflow for a pea or canola panel.
:::

## 2. How many groups? Clustering with validation

```{r}
library(mclust)
km  <- kmeans(p$x[,1:5], centers = 3, nstart = 25)
mc  <- Mclust(p$x[,1:5], G = 1:6)     # model-based; picks G by BIC
mc$G
fviz_cluster(km, data = p$x[,1:2], geom = "point") + theme_minimal()
```

k-means needs k chosen; `mclust` fits Gaussian mixtures and lets **BIC** choose. Report silhouette or BIC, never just a pretty plot.

## 3. Supervised: sparse PLS-DA

PCA ignores labels. **PLS-DA** finds components that separate known groups; the **sparse** version selects a small marker/metabolite panel — interpretable and directly useful for assay design.

```{r}
library(mixOmics)
Y <- factor(km$cluster)                     # demo labels; use market class/origin in practice
res <- splsda(M, Y, ncomp = 2, keepX = c(30, 20))
plotIndiv(res, ellipse = TRUE, legend = TRUE, title = "sPLS-DA")
```

**Honest validation — mandatory:**

```{r}
#| eval: false
perf(res, validation = "Mfold", folds = 5, nrepeat = 20)   # error rates
# + permutation test: shuffle Y, refit — separation should collapse
```

sPLS-DA will happily "separate" random labels; permutation testing is what keeps you honest (a favourite interview probe).

## 4. GxE with PCA: the AMMI idea

```{r}
#| fig-height: 3.5
library(agridat); data(gauch.soy)
gxe <- gauch.soy |> group_by(gen, env) |> summarise(y = mean(yield), .groups="drop") |>
  pivot_wider(names_from = env, values_from = y) |> column_to_rownames("gen")
gxe_c <- sweep(sweep(as.matrix(gxe), 1, rowMeans(gxe)), 2, colMeans(gxe)) + mean(as.matrix(gxe))
pg <- prcomp(gxe_c)
fviz_pca_biplot(pg, repel = TRUE, geom = "point", title = "AMMI-style GxE biplot")
```

Genotypes near the origin are stable; those aligned with an environment vector are specifically adapted — the statistical heart of MET interpretation.

## Exercises

1. Choose k for the SNP data by silhouette width; does it agree with mclust's BIC?
2. Run `perf()` on the sPLS-DA — report the balanced error rate per component.
3. Replace demo labels with `DT_cpdata$color` classes and re-select markers; how many are needed for <10% error?

**Data credits:** `DT_cpdata` (Covarrubias-Pazaran, sommer); Gauch soybean (agridat).
