---
title: "Week 7 — Databases & APIs for Breeding Data"
subtitle: "From flat files to a queryable, serveable resource"
---

```{mermaid}
flowchart LR
    A[Relational model<br/>lines · markers · phenotypes] --> B[SQLite<br/>create & load]
    B --> C[SQL verbs<br/>SELECT · JOIN · GROUP BY]
    C --> D[dbplyr<br/>dplyr → SQL]
    D --> E[plumber API<br/>GET /gwas?trait=...]
```

## 1. Why a database beats 40 CSVs

One authoritative copy, enforced types and keys, concurrent access, and queries that don't require loading 2 GB into RAM. The mental model: **one table per entity**, keys linking them.

```
lines(line_id, name, family, origin)
phenotypes(line_id, trait, env, value)
gwas_results(trait, marker, chr, pos, pvalue)
```

## 2. Build it in R (fully reproducible)

```{r}
library(DBI); library(RSQLite); library(dplyr)
con <- dbConnect(SQLite(), "course.db")

set.seed(3)
lines <- tibble(line_id = 1:200, name = paste0("L", 1:200),
                family = sample(paste0("F", 1:10), 200, TRUE))
phenos <- tidyr::expand_grid(line_id = 1:200, trait = c("Yield","Oil"),
                             env = c("Saskatoon","Melfort")) |>
  mutate(value = rnorm(n(), 50, 8))
dbWriteTable(con, "lines", lines, overwrite = TRUE)
dbWriteTable(con, "phenotypes", phenos, overwrite = TRUE)
dbListTables(con)
```

## 3. SQL you'll actually write

```{r}
dbGetQuery(con, "
  SELECT l.family, p.trait, ROUND(AVG(p.value),1) AS mean_val, COUNT(*) AS n
  FROM phenotypes p JOIN lines l USING(line_id)
  WHERE p.env = 'Saskatoon'
  GROUP BY l.family, p.trait
  ORDER BY mean_val DESC LIMIT 6")
```

## 4. Or let dbplyr write the SQL

```{r}
tbl(con, "phenotypes") |>
  left_join(tbl(con, "lines"), by = "line_id") |>
  group_by(family, trait) |>
  summarise(mean_val = mean(value), .groups = "drop") |>
  show_query()
```

Same dplyr verbs as Week 0b — translated to SQL and executed *in* the database.

```{r}
#| include: false
dbDisconnect(con)
```

## 5. Serve it: a REST API in 20 lines

`api/plumber.R`:

```{r}
#| eval: false
library(plumber); library(DBI); library(RSQLite)

#* @apiTitle Breeding data API — Sampath Perumal

#* Mean phenotype by family
#* @param trait Trait name
#* @get /summary
function(trait = "Yield") {
  con <- dbConnect(SQLite(), "course.db"); on.exit(dbDisconnect(con))
  dbGetQuery(con, "SELECT l.family, AVG(p.value) AS mean_val
                   FROM phenotypes p JOIN lines l USING(line_id)
                   WHERE p.trait = ? GROUP BY l.family", params = list(trait))
}
```

Run: `plumber::plumb("api/plumber.R")$run(port = 8000)` → `http://localhost:8000/summary?trait=Oil` returns JSON. Your Week 6 Shiny app, a Python script, and a collaborator's browser all consume the **same endpoint** — that's the architecture sentence worth saying in interviews.

## Exercises

1. Add a `gwas_results` table from Week 3/4 output; write a query returning the top 10 markers per trait (window function: `ROW_NUMBER() OVER (PARTITION BY trait ORDER BY pvalue)`).
2. Add a `/gwas` endpoint with a p-value filter parameter.
3. Point the Week 6 app at the database via `pool` + `DBI` instead of RDS files.
