---
title: "Week 0b — R Essentials"
subtitle: "Just enough R to be dangerous, in one sitting"
---

```{mermaid}
flowchart LR
    A[Projects &<br/>working directory] --> B[Vectors &<br/>data frames]
    B --> C[tidyverse verbs<br/>filter · mutate · summarise]
    C --> D[ggplot2<br/>grammar of graphics]
    D --> E[Functions &<br/>reading genomic files]
    E --> F[Quarto<br/>reports]
```

## Projects and paths

Always work inside an **RStudio Project** (`.Rproj`) and build paths with `here::here()` — scripts then run on any machine, which is half of reproducibility.

## Vectors and data frames

```{r}
yield  <- c(42.1, 39.8, 45.3, 41.0)          # numeric vector
line   <- c("L1", "L2", "L3", "L4")
trial  <- data.frame(line, yield)
trial$yield_kg <- trial$yield * 67.25         # bu/ac -> kg/ha (canola)
trial
```

## The six tidyverse verbs that do 90% of work

```{r}
library(tidyverse)
data(oats, package = "agridat")   # a real 1930s split-plot oat trial (Yates)
glimpse(oats)

oats |>
  filter(nitro > 0) |>
  mutate(yield_t = yield * 0.0251) |>
  group_by(gen, nitro) |>
  summarise(mean_yield = mean(yield), .groups = "drop") |>
  arrange(desc(mean_yield)) |>
  head()
```

`filter` rows · `select` columns · `mutate` new variables · `group_by`+`summarise` aggregate · `arrange` sort · `left_join` merge (genotype ↔ phenotype tables).

## ggplot2: plots as grammar

```{r}
#| fig-height: 3.5
ggplot(oats, aes(x = factor(nitro), y = yield, fill = gen)) +
  geom_boxplot() +
  labs(x = "Nitrogen (cwt/ac)", y = "Yield", fill = "Genotype",
       title = "Yates (1935) oat trial") +
  theme_minimal()
```

Aesthetics map data → visual channels; geoms draw; facets split. Every figure in this course is built this way.

## Reading genomic-style files

```{r}
#| eval: false
pheno <- read_csv("data/phenotypes.csv")
geno  <- read_tsv("data/genotypes.hmp.txt")          # HapMap
vcf   <- vcfR::read.vcfR("data/snps.vcf.gz")         # VCF
```

## Writing functions

```{r}
cv <- function(x) 100 * sd(x, na.rm = TRUE) / mean(x, na.rm = TRUE)
cv(oats$yield)
```

If you paste code three times, it should be a function; if a function is used in three scripts, it belongs in `R/helpers.R`.

## Quarto

This entire website is Quarto: prose + chunks + output in one `.qmd`. Render with the **Render** button or `quarto render`. Your analyses become reports automatically — this is the reporting standard the course assumes.

## Exercises

1. With `agridat::oats`, compute the CV of yield per genotype.
2. Plot yield vs nitrogen as points + a loess smooth, faceted by genotype.
3. Write `standardize <- function(x)` returning z-scores; apply it to yield within each block using `group_by() |> mutate()`.
